Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-05-19 doc/manuscript/translocation.qmd
Revise Abstract based on comments from coauthor EBR

Escalating anthropogenic change is devastating global biodiversity.
To avoid extinction, affected populations will often need to adapt to the altered environment.
The process of rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth is termed "evolutionary rescue".
In addition to allowing recovery of extant populations, individuals from rescued populations could also be used to reestablish extirpated populations.
We evaluated this scenario in the once-common mountain yellow-legged (MYL) frog, which is threatened with extinction by the human-facilitated emergence of a lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd").
Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually begin to recover.
We show that MYL frogs have undergone substantial evolutionary change following disease outbreaks, and that these changes are associated with increased resistance/tolerance to Bd infection in naturally recovering populations.
Large-scale reintroduction of frogs from these populations resulted in the establishment of reproducing populations at most recipient sites despite ongoing disease.
In addition, results from viability modeling suggest that established populations have a low probability of extinction over 50 years.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant/tolerant individuals can allow the landscape-scale recovery of disease-impacted species.
This example has potentially broad implications for the many taxa worldwide that are threatened with extinction by anthropogenic stressors.

@@

Human activities are intensifying threats to global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
To persist, affected taxa will often need to adapt to their altered environment or face decline and possible extinction.
The process whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth is known as "evolutionary rescue" [@carlson2014].
In addition to preventing extinction of declining populations, individuals from rescued populations could also disperse, or be reintroduced, to empty habitats and promote larger-scale recovery.
Although this overall scenario is potentially of critical importance for the conservation of an ever-growing number of threatened taxa, descriptions of its occurrence and effectiveness in the wild are lacking.

@@

Evolutionary rescue may be increasingly critical for the survival of wildlife populations in a rapidly changing world.
The importance of evolutionary rescue is known for a wide variety of taxa [e.g., @carlson2014; @mills2018; @oziolor2019], including in the specific context of disease-impacted systems [@searle2020].
However, obtaining evidence in wild animal populations for all the causal links in this process (i.e., selection for adaptive alleles $\rightarrow$ increase in host fitness in the presence of the stressor $\rightarrow$ positive population growth and recovery) is difficult in most systems.
Nonetheless, we provide a compelling example from the wild not only that evolutionary rescue can produce individuals that harbor adaptive alleles and allow population recovery, but importantly that these individuals can be used in translocations to reestablish extirpated populations.
We expect that such evolutionary rescue-based translocations will be an essential tool in wildlife conservation in an era of accelerating global change.<!--# Dig deeper into the "difficult" task alluded to above? E.g., what would it take to increase confidence that evolutionary rescue is at play, what alternative explanations are still credible given the available data? -->

@@

We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
The possible states of individuals include "not recruited", "alive", and "dead".
The possible observations of individuals include "detected" and "not detected".
We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
Code used to create CMR datasets and fit models for each translocated population is available from the following GitHub repository: <https://github.com/SNARL1/cmr-analysis>.

<!--# Add text describing calculations of VES and CMR-recruitment counts. For the latter, only included surveys conducted >= 3 years following translocation to account for duration of early life stages. -->

@@

We conducted two analyses.
In the first, we addressed the more general question of whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
Group-level effects included site_id or translocation_id.
The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load (bd_load), sex (sex_male), frog size (length), site elevation (elevation), winter severity in the year of translocation (snow_t), winter severity in the year following translocation (snow_t1), donor population (donor_72996), day of year on which a translocation was conducted (day), and translocation order (order_first)).
In both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3--16).
Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
In both analyses, the response variable, probability of 1-year frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
For all models, we used default priors, four chains, and 5000 iterations each for warmup and post-warmup.
We checked convergence of the MCMC algorithm using trace plots and Rhat.
Code to create the datasets and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.

@@

We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see SI Appendix for details].
To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
Before-translocation loads were quantified using skin swabs collected from all to-be-translocated frogs at the donor site on the day before or the day of the translocation.
After-translocation Bd loads were based on all swabs collected from translocated frogs at the recipient site in the year of and the year following translocation.
Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the year 0-1 period.
We used an analysis workflow similar to that described above.
Briefly, the response variable was before-translocation load and the predictor variable was after-translocation load.
We compared models with and without group-level effects, and group-level effects included pit_tag_id, pit_tag_id nested within site_id, or pit_tag_id nested within translocation_id.
All models used a negative binomial family.
Priors, chains, iterations, and checks were as described above.
Code to create the dataset and fit models is available from the following GitHub repository: <https://github.com/SNARL1/translocation>.
<!--# Does it make more sense to only analyze the model with pit_tag_id as the group-level effect? That is the model that is most analogous to a paired t-test. -->

@@

In this model we ignore density-dependent recruitment because we were interested in the growth of the population from an initial reintroduction and whether this growth was sufficient to prevent extinction over 50 years following the introduction.
We also did not directly consider the dynamics of Bd in this model.
We made this decision because (i) translocated populations are infected with Bd at high prevalence [@joseph2018], and (ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022].
Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
(Additional details regarding incorporating yearly variability in vital rates and estimating model parameters are provided in the SI Appendix.)

@@

After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses (described in detail in the SI Appendix).
First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
Third, we included demographic and environmental stochasticity in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
<!--# Add link to Mark's GitHub repo? -->

@@

Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
We focused on population 70550 because this was our longest CMR time series for a translocated population and because this population shows evidence of substantial post-translocation increases in adult abundance associated with population establishment and recovery.
We simulated our model for 16 years, repeating the simulation 50,000 times.
For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.

>>>>

Vast alteration of the biosphere by humans is causing a sixth mass extinction.
Adaptation to modified environments often is the only means for species to persist, but the extent to which such "evolutionary rescue" can prevent or reverse biodiversity loss is largely unknown.
Using results from genomic analyses, a 15-year reintroduction effort, and population modeling, we provide a compelling example of evolution reversing the declines of an imperiled amphibian and allowing the reestablishment of extirpated populations.
The once-common mountain yellow-legged (MYL) frog is threatened with extinction by the human-facilitated emergence of a lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd").
Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually begin to recover.
Using an exome capture approach, we show that MYL frogs have undergone substantial evolutionary change following disease outbreaks.
Moreover, in naturally recovering populations observed changes in immune function loci are associated with increased resistance/tolerance to Bd infection.
Large-scale reintroduction of frogs from rescued populations resulted in the establishment of reproducing populations despite ongoing disease.
In addition, results from viability modeling suggest that established populations have a low probability of extinction over 50 years.
Collectively, these results provide a rare example of how evolutionary rescue and the reintroduction of resistant/tolerant individuals can allow the landscape-scale recovery of disease-impacted species.
This example has potentially broad implications for the many taxa worldwide that are threatened with extinction by anthropogenic stressors.

@@

Human activities are intensifying threats to global biodiversity [@ceballos2015], with important implications for ecosystem resilience and human welfare [@naeem2009].
To persist, affected taxa will often need to adapt to their altered environment or face decline and possible extinction.
The process whereby rapid evolutionary change increases the frequency of adaptive alleles and restores positive population growth is known as "evolutionary rescue" [@carlson2014].
In addition to preventing extinction of declining populations, individuals from rescued populations could also disperse, or be reintroduced, to empty habitats and promote larger-scale recovery.
Although this overall scenario is potentially of critical importance for the conservation of an ever-growing number of imperiled taxa, descriptions of its occurrence and effectiveness in the wild are lacking.

@@

Evolutionary rescue may be increasingly critical for the survival of wildlife populations in a rapidly changing world.
The importance of evolutionary rescue is known for a wide variety of taxa [e.g., @carlson2014; @mills2018; @oziolor2019], including in the specific context of disease-impacted systems [@searle2020].
However, obtaining evidence in wild animal populations for all the causal links in this process (i.e., selection for adaptive alleles $\rightarrow$ increase in host fitness in the presence of the stressor $\rightarrow$ positive population growth and recovery) is difficult in most systems.
Nonetheless, we provide a compelling example from the wild not only that evolutionary rescue can produce individuals that harbor adaptive alleles and allow population recovery, but importantly that these individuals can be used in translocations to reestablish extirpated populations.
We expect that such evolutionary rescue-based translocations will be an essential tool in wildlife conservation in an era of accelerating global change.

@@

We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and adult population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions to the population (i.e., translocated cohorts), as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4, @rsoftware2022].
The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
The possible states of individuals include "not recruited", "alive", and "dead".
The possible observations of individuals include "detected" and "not detected".
We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).

@@

We conducted two analyses.
In the first, we addressed the more general question of whether survival of frogs in the first translocation to a site predicted the survival of frogs in subsequent translocations.
Group-level effects included site_id or translocation_id.
The second analysis focused specifically on identifying predictors of post-translocation frog survival using a wide range of variables describing characteristics of frogs, sites, and translocations (predictor variables: Bd load (bd_load), sex (sex_male), frog size (length), site elevation (elevation), winter severity in the year of translocation (snow_t), winter severity in the year following translocation (snow_t1), donor population (donor_72996), day of year on which a translocation was conducted (day), and translocation order (order_first)).
In both analyses we used 1-year survival as a single metric of survival that was available for all translocated populations regardless of how many years of post-translocation data were available (range = 3--16).
Group-level effects included site_id, translocation_id, or translocation_id nested within site_id.
In both analyses, the response variable, probability of 1-year frog survival, was rounded to integer values to produce a binomial outcome, and all models used a bernoulli family.
For all models, we used default priors, four chains, and 5000 iterations each for warmup and post-warmup.
We checked convergence of the MCMC algorithm using trace plots and Rhat.

@@

We analyzed skin swabs using standard Bd DNA extraction and qPCR methods [@boyle2004, see SI Appendix for details].
To assess the magnitude of changes in Bd load on frogs following translocation, we compared Bd loads measured before versus after translocation.
Before-translocation loads were quantified using skin swabs collected from all to-be-translocated frogs at the donor site on the day before or the day of the translocation.
After-translocation Bd loads were based on all swabs collected from translocated frogs at the recipient site in the year of and the year following translocation.
Individual frogs and their associated Bd loads were included in the dataset only if frogs were captured at the recipient site at least once during the year 0-1 period.
We used an analysis workflow similar to that described above.
Briefly, the response variable was before-translocation load and the predictor variable was after-translocation load.
We compared models with and without group-level effects, and group-level effects included pit_tag_id, pit_tag_id nested within site_id, or pit_tag_id nested within translocation_id.
All models used a negative binomial family.
Priors, chains, iterations, and checks were as described above.

Code to replicate all of the analyses described above is available from the following GitHub repositories: <https://github.com/SNARL1/translocation>; <https://github.com/SNARL1/cmr-analysis>.

@@

In this model we ignore density-dependent recruitment because we were interested in the growth of the population from an initial reintroduction and whether this growth was sufficient to prevent extinction over 50 years following the introduction.
We also did not directly consider the dynamics of Bd in this model.
We made this decision because (i) translocated populations are infected with Bd at high prevalence [@joseph2018], and (ii) host density does not seem to play a significant role in multi-year Bd infection dynamics in this system [@wilber2022].
Thus, ignoring Bd infection dynamics and instead assuming all host vital rates are in the presence of high Bd prevalence significantly simplifies the model without much loss of realism.
Additional details are provided in the SI Appendix.

@@

After parameterizing our model with CMR-estimated adult frog survival probabilities and other known vital rates ([@tbl-param_values]), we performed four analyses (described in detail in the SI Appendix).
First, we compute the long-run growth rate $\lambda$ for each of our 12 translocated populations to determine if the populations were deterministically predicted to grow or decline in the long-run.
Second, we compute the elasticity of $\lambda$ to model parameters, to quantify how much changes in adult survival, for example, affected the long-run growth rate.
Third, we included demographic and environmental stochasticity in our model and simulated the 50-year viability (i.e., 1 - extinction probability) of populations given an introduction of 40 adult individuals into an unoccupied habitat.
Finally, we fit our model to our longest translocation trajectory to confirm that our model could reasonably reproduce the observed recovery trajectories of MYL frogs following reintroductions.
Additional details are provided in the SI Appendix.
Code to replicate the analyses can be found at <https://github.com/SNARL1/translocation>.

@@

Finally, we assessed whether our stochastic model could reproduce observed trajectories of population recovery.
We focused on population 70550 because this was our longest CMR time series for a translocated population and because this population shows evidence of substantial post-translocation increases in adult abundance associated with population establishment and recovery.
We simulated our model for 16 years, repeating the simulation 50,000 times.
For each run and each year, we drew $\omega$ from a uniform distribution between 0 and 1 (or equivalently a beta distribution with mean 0.5 and $\phi = 2$).
Using Approximate Bayesian Computing and rejection sampling [@kosmala2016], we identified the top 2% of trajectories (i.e., 1000 trajectories) that minimized the sum of squared errors between the observed and predicted data.
The $\omega$ values associated with these "best" trajectories represented an approximate posterior distribution [@beaumont2010].
Using these best fit trajectories, we assessed whether our model could qualitatively describe the patterns of recovery in the observed data for population 70550.
